---
title: "Constrained Longitudinal Data Analysis (cLDA) in R"
author: "Alexandre Bohyn"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{cLDA}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE,
  dpi = 300
)
```

```{r setup}
# Packages used in the analysis
library(nlme) # GLS + correlation / variance structures
library(marginaleffects) # avg_predictions(), hypotheses()
# library(ggplot2)         # plotting
library(tinyplot)
```

```{r}
# Read the example data from the package
data(priority, package = "lbstfun")
```

## Custom design matrix

To fit a cLDA model, we need to create a custom design matrix that includes the time variable and its interaction with the treatment group.
It should not contain the intercept, nor the interaction with the first level of time (baseline).
We also relabel the columns to avoid issues with special characters (e.g., `:`).

```{r}
# Create the design matrix (no ME for group)
design_matrix_full <- model.matrix(~ time + time:trt, data = priority)

# Remove the intercept and the interaction with the first level of time (baseline)
col_names <- colnames(design_matrix_full)
selected_cols <- !(col_names %in% c("(Intercept)", "time0:trt2"))
design_matrix <- design_matrix_full[, selected_cols]

# Relabel the columns to avoid issues with special characters (:)
relabelled_colnames <- gsub(":", "_", col_names[selected_cols])
colnames(design_matrix) <- relabelled_colnames

# Combine the design matrix with the outcome and id
model_data <- cbind(
  priority,
  as.data.frame(design_matrix)
)

# View the first rows of the custom design matrix
head(design_matrix)
```

## Fit the model

Since the design matrix was created manually, we need to specify all the components of the model ourselves in the formula.
We also specify the variance structure (one variance per time point) and the correlation structure (unstructured correlation within subjects).

```{r}
# Create the formula for the model
form <- formula(paste(
  "outcome ~",
  paste(colnames(design_matrix), collapse = " + ")
))

# Fit the constrained LDA model
fit <- gls(
  form,
  data = model_data,
  weights = varIdent(form = ~ 1 | time), # One variance per time point
  correlation = corSymm(form = ~ 1 | id) # Unstructured correlation within subjects
)

coef(summary(fit))
```

## Estimated marginal means and contrasts

We first compute the predicted values at each time point for each treatment group and plot them.

```{r}
# Obtain average predictions at each time point for each treatment group
p <- avg_predictions(fit, by = c("time", "trt"))
p
```

```{r}
#| fig.width: 7
#| fig.height: 4
#| dev: "svg"

# Plot these predictions
plt(
  estimate ~ time | trt,
  data = p,
  ymin = conf.low,
  ymax = conf.high,
  type = type_lines(dodge = 0.05),
  col = c("#0050ef", "#e51400"),
  legend = list(title = "Treatement"),
  xlab = "Time",
  ylab = "Estimate"
)
plt_add(type = type_errorbar(dodge = 0.05))
```

To test specific hypotheses, we can use the `hypotheses()` function.
We then specify which linear combination of the predicted values in `p` we want to test.
For example, `"b5 - b1"` tests the difference between row 5 and row 1 of the predicted values (i.e., time 2 vs time 0 for treatment group 1).

```{r}
# Compute the difference between time 2 and time 0 for each treatment group
hypotheses(p, hypothesis = "b5 - b1 = 0") # For the reference group (trt1)
hypotheses(p, hypothesis = "b6 - b2 = 0") # For the other group (trt2)
```

We can also test if the change over time is different between the two treatment groups (i.e., difference in differences).

```{r}
# Compute the difference in differences between groups at time 2 vs time 0
hypotheses(p, hypothesis = "(b6 - b2) = (b5 - b1)")
```

## Covariance matrix

```{r}
# Extract correlation matrix (1 per subject )
cor_matrix <- corMatrix(fit$modelStruct$corStruct, corr = TRUE)

# Extract variance weights (relative to base level)
var_weights <- coef(
  fit$modelStruct$varStruct,
  unconstrained = FALSE,
  allCoef = TRUE
)

# Get base residual variance
base_var <- fit$sigma^2

# Compute actual variances per time level
var <- base_var * var_weights^2
sd <- sqrt(var)

# Build covariance matrices for first subject
cor <- cor_matrix[[1]]
cov <- diag(sd) %*% cor %*% diag(sd)

cov
```

