---
title: "Interrupted Time Series (ITS) in R"
author: "Alexandre Bohyn"
output: 
  rmarkdown::html_vignette:
    code_folding: "show"
vignette: >
  %\VignetteIndexEntry{ITS}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE
)
```

```{r setup}
# Packages used in the analysis
library(marginaleffects) # predictions(), slopes(), hypotheses()
library(ggplot2) # plotting
library(cowplot) # plotting theme
library(see) # color palette
library(dplyr) # data manipulation
library(tidyr) # uncount()
library(forcats) # fct_cross()
```

## Data generation

In an interrupted time series design, outcomes are measured repeatedly over time for two groups (here, Control and Exposed), and a common intervention occurs partway through the observation period.
We generate a synthetic dataset with 6 time points and an intervention between time points 3 and 4.
Ten subjects are simulated per time-by-group cell, with outcomes drawn from a normal distribution around known group means.

The key derived variable is `grp`, a four-level factor created by crossing the intervention period (Pre / Post) with the exposure status (Control / Exposed).
This encodes the four distinct trajectories in the data: `Pre_Control`, `Pre_Exposed`, `Post_Control`, and `Post_Exposed`.

```{r data-creation}
# Generate a synthetic ITS dataset
df <- expand.grid(
  time = 1:6,
  exposed = 0:1
) |>
  mutate(mean = c(26, 30, 32, 45, 52, 63, 27, 31, 33, 29, 32, 36)) |>
  uncount(weights = 10, .id = "id") |>
  mutate(
    outcome = mean + rnorm(n = n(), sd = 3),
    intervention = as.numeric(time > 3) |>
      factor(levels = c(0, 1), labels = c("Pre", "Post")),
    exposed = factor(
      exposed,
      levels = c(0, 1),
      labels = c("Control", "Exposed")
    ),
    grp = fct_cross(intervention, exposed, sep = "_") # 4-level group variable
  )
```

## Exploratory plot

Before fitting a model, it is useful to visualize the raw data alongside group means to inspect the trajectory of each group across time and the apparent effect of the intervention.
The dashed vertical line marks the intervention point (between time 3 and time 4).

```{r data-plot}
#| class.source: "fold-hide"
#| fig.width: 6

df |>
  mutate(exp_time = paste0(exposed, "_", time)) |>
  ggplot(aes(x = time, y = outcome, color = exposed)) +
  geom_boxplot(aes(group = exp_time), fill = NA) +
  geom_vline(xintercept = 3.5, linetype = "dashed") +
  geom_point(aes(group = exp_time), position = position_dodge(width = 0.8)) +
  stat_summary(
    aes(group = grp),
    geom = "line",
    fun = mean,
    position = position_dodge(width = 0.8)
  ) +
  scale_x_continuous(breaks = 1:6) +
  see::scale_color_material(name = "Exposure") +
  labs(x = "Time", y = "Outcome") +
  cowplot::theme_cowplot(12)
```

## Fit the model

We fit a linear model that allows each group to have its own intercept and its own time slope.
Removing the global intercept (`- 1`) means the four levels of `grp` directly estimate the group-specific baseline values.
The `grp:time` interaction then gives each group its own rate of change over time, so the model can capture both a level shift and a slope change at the intervention.

```{r}
# Group-specific intercepts and time slopes; no global intercept
model <- lm(outcome ~ grp + grp:time - 1, data = df)
parameters::parameters(model)
```

## Estimated marginal means

We compute the predicted outcome at each time point for each group combination.
These marginal means integrate out subject-level variability and provide a clean summary of the fitted trajectories.

```{r}
# Average predicted values at each time × group combination
preds <- predictions(
  model,
  by = c("time", "grp")
)
preds
```

## Level change at intervention

The level change captures the immediate shift in outcome at the intervention boundary, defined here as the difference between time 3 (last pre-intervention time point) and time 4 (first post-intervention time point) within each group.
We build contrast vectors by logical indexing into `preds`, then pass them to `hypotheses()` to test three quantities simultaneously: the level change in the Control group, the level change in the Exposed group, and the difference between the two (i.e., whether the intervention had a differential effect on level).

```{r}
# Contrast vectors: pre-intervention endpoint minus post-intervention starting point
ctrl <- with(preds, time == 3 & grp == "Pre_Control") -
  with(preds, time == 4 & grp == "Post_Control")
exp <- with(preds, time == 3 & grp == "Pre_Exposed") -
  with(preds, time == 4 & grp == "Post_Exposed")

# Test level changes and their difference
hypotheses(
  preds,
  hypothesis = cbind(
    "Control" = ctrl,
    "Exposed" = exp,
    "Difference" = exp - ctrl
  )
)
```

## Trend analysis

Beyond a level shift, the intervention may also alter the rate of change over time (i.e., the slope).
We use `slopes()` to estimate the time trend within each group-period combination, yielding four slope estimates (Pre_Control, Post_Control, Pre_Exposed, Post_Exposed).
We then use `hypotheses()` to test three contrasts: the change in slope from Pre to Post for the Control group, the same change for the Exposed group, and the difference-in-slopes between the two groups (an estimate of the differential trend effect of the intervention).

```{r}
# Estimate time slopes within each group × period combination
trends <- slopes(
  model,
  variables = "time",
  by = "grp"
)
trends
```

```{r}
# Test the change in trend from Pre to Post, and the difference between groups
hypotheses(
  trends,
  hypothesis = c(
    "Control" = "b2 - b1 = 0",
    "Exposed" = "b4 - b3 = 0",
    "Difference" = "(b4 - b3) - (b2 - b1) = 0"
  )
)
```

